# Summary Statistics

Using the `psych::describe()` function


```{r, message=FALSE, error=FALSE}
library(psych)           # Lots of good tidbits
```


The `describe()` function from the `psych` package returns an extensive listing of **basic summary statistics** for every variable in a dataset [@R-psych]. 

* `vars` number order of the variables in this table
* `n` how many non-missing values there are
* `mean` the average or arithmetic mean
* `sd` the standard deviation 
* `median` the 50th percentile or Q2
* `trimmed` the mean after removing the top and bottom 10% of values
* `mad` median absolute deviation (from the median) DO NOT WORRY ABOUT!
* `min` the minimum or lowest value
* `max` the maximum or highest value
* `range` full range of values, max - min
* `skew` skewness (no SE for skewness given)
* `kurtosis` kurtosis (no SE for kurtosis given)
* `se` the standard error for the **MEAN**, not the skewness or kurtosis


## All Variables in a Dataset


```{r}
cancer_clean %>% 
  psych::describe()
```

**NOTE** The names of categorical variables *(factors)* are followed by an **astrics** to indicate that summary statistics should not be evaluated since the variable is not continuous or on an interval scale.


## A Subset of Varaibles in a Datasets

It is better to avoid calculating summary statistics for categorical variables in the first place by first restricting the dataset to only continuous variables using a `dplyr::select()` step.

> Make sure to use a `dplyr::select(var1, var2, ..., var12)` step to select only the variables of interest.

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, totalcin, totalcw2, totalcw4, totalcw6) %>%
  psych::describe()
```

## Turn off reporting Skewness adn Kurtosis

```{r}
cancer_clean %>% 
  dplyr::select(age, weighin, totalcin, totalcw2, totalcw4, totalcw6) %>%
  psych::describe(skew = FALSE)
```




