# Simple and Stratified Descriptive Tables


Using the `furniture::table1()` function

```{r, message=FALSE, error=FALSE}
library(furniture)       # nice tables of descriptives
```


The `table1()` function in the `furniture` package returns a much smaller listing of summary statistics [@R-furniture].

* Categorical Variables: count (percentage) within each category
* Continuous Variables: mean (standard deviation)



## For a Single Categorical Variable

```{r}
data_ihno %>% 
  furniture::table1(majorF)
```

## For a Single Continuous Variable

```{r}
data_ihno %>% 
  furniture::table1(statquiz)
```


## For a Subset of Variables


The variables desired can also be listed inside the function *(selarate with commas)*.



```{r}
cancer_clean %>% 
  furniture::table1(age, weighin, stage, totalcin)
```




```{r}
data_ihno %>% 
  furniture::table1(majorF, reasonF, statquiz)
```




## Customize the Variable Labels

```{r}
cancer_clean %>% 
  furniture::table1("Age, years"     = age, 
                    "Weight, pounds" = weighin, 
                    "Cancer Stage"   = stage, 
                    "Oral Condition" = totalcin)
```



```{r}
data_ihno %>% 
  furniture::table1("Reason For Enrolling" = reasonF, 
                    "Coffee Drinker"       = coffeeF, 
                    "Math Phobia"          = phobia, 
                    "Statistics Quiz"      = statquiz, 
                    "Mathematics Quiz"     = mathquiz) 
```

## Chance rounding 

```{r}
cancer_clean %>% 
  furniture::table1(age, weighin, stage, totalcin,
                    digits = 2)
```
```{r}
cancer_clean %>% 
  furniture::table1(age, weighin, stage, totalcin,
                    digits = 3)
```

## Stratify -by- a categorical factor


You may use a `dplyr::group_by(grouping_var)` step before the `furniture::table1()` step to create summary statistics for different subgroups. 


```{r}
cancer_clean %>% 
  dplyr::group_by(trt) %>% 
  furniture::table1("Age, years"     = age, 
                    "Weight, pounds" = weighin, 
                    "Cancer Stage"   = stage, 
                    "Oral Condition" = totalcin)
```



```{r}
data_ihno %>% 
  dplyr::group_by("Gender" = genderF) %>% 
  furniture::table1("College Major" = majorF, 
                    "Reason Enrolled" = reasonF,
                    "Statistics Quiz" = statquiz)
```



```{r}
data_ihno %>% 
  dplyr::mutate(phobia_cut3 = cut(phobia, 
                                  breaks = c(0, 2, 4, 10),
                                  include.lowest = TRUE)) %>% 
  dplyr::group_by("Self-reported Math Phobia Rating" = phobia_cut3) %>% 
  furniture::table1("College Major" = majorF, 
                    "Reason Enrolled" = reasonF,
                    "Statistics Quiz" = statquiz)
```




## Add a TOTAL Column

```{r}
cancer_clean %>% 
  dplyr::group_by(trt) %>% 
  furniture::table1("Age, years"     = age, 
                    "Weight, pounds" = weighin, 
                    "Cancer Stage"   = stage, 
                    "Oral Condition" = totalcin,
                    total = TRUE)
```

```{r}
data_ihno %>% 
  dplyr::group_by("Gender" = genderF) %>% 
  furniture::table1("College Major" = majorF, 
                    "Reason Enrolled" = reasonF,
                    "Statistics Quiz" = statquiz,
                    total = TRUE,
                    digits = 2)
```

