# Boxplots


Using the `ggplot2::geom_boxplot()` function

```{r, message=FALSE, error=FALSE}
library(tidyverse)       # super helpful everything, including ggplot
```

For boxplots, you must specify the variable of interest within the aesthetics as the `y` variable: `ggplot(aes(y = continuous_var))` before adding the `geom_boxplot()` layer [@R-ggplot2].

> Reminder: Steps BEFORE the `ggplot()` are combined with pipes `%>%`, whereas layers of the plot are combined with the addition symbol `+`.

## Single continuous variable

If you only want to produce a single boxplot, then the aesthetics must include **some quoted text** as the `x` variable.

```{r}
cancer_clean %>%
  ggplot(aes(x = "Full Data Set", 
             y = weighin)) +
  geom_boxplot()
```


## Single Box -for- a Subset

### One Requirement 

Where as the `dplyr::select()` function specifies which VARIABLES to reduce down to, the `dplyr::filter()` function specifies which ROWS or PARTICIPANTS to reduce down to.

> When using a `dplyr::filter()` step, make sure to change the `x = "text"`, too.

```{r}
cancer_clean %>%
  dplyr::filter(weighin < 172) %>%
  ggplot(aes(x = "Weigh At Baseline < 172", 
             y = weighin)) +
  geom_boxplot()
```

### Two Requirements 

When testing for a match to a value *(for equality)* use `==` instead of `=` in the `dplyr::filter()` step.

Use the `&` symbol to require multiple conditions for the subset, but only include one quoted text phrase for `x`.

```{r}
cancer_clean %>%
  dplyr::filter(weighin >= 150 & trt == "Placebo") %>%
  ggplot(aes(x = "Placebo and at least 150 Pounds", 
             y = weighin)) +
  geom_boxplot()
```


### A Requirement Specified with a List

A helpful symbol-set is `%in%`, which tests if the thing *before* it is **included in** the  list of elements that comes *after* it.

```{r}
cancer_clean %>%
  dplyr::filter(trt == "Aloe Juice" & stage %in% c(2, 3, 4)) %>%
  ggplot(aes(x = "On Aloe Juice and Stage 2-4", 
             y = weighin)) +
  geom_boxplot()
```



## Multiple Boxes -by- a Factor

There are **three** ways to included a categorical variable to break the sample into groups. You may specify the factor with:

(@) `fill` the variable is denoted with different colors filling the boxes
(@) `x` the horizontal axis is marked for all levels
(@) `facet_grid()` a panel is create for each level


### Use `fill = var_name` 

```{r}
cancer_clean %>%
  ggplot(aes(x = "Full Sample",
             y = weighin,
             fill = trt)) +
  geom_boxplot()
```

### Use `x = var_name` 

```{r}
cancer_clean %>%
  ggplot(aes(x = trt, 
             y = weighin)) +
  geom_boxplot()
```


### Use `facet_grid(. ~ var_name)` 


```{r}
cancer_clean %>%
  ggplot(aes(x = "Full Sample",
             y = weighin)) +
  geom_boxplot() +
  facet_grid(. ~ trt)
```

## Multiple Boxes -by- 2 Factors

You may combine any of the three previous specification, one per categorical (`factor`) variable.

### Use `fill = var_name_1` and `x = var_name_2`

```{r}
cancer_clean %>%
  ggplot(aes(x = stage,
             y = weighin,
             fill = trt)) +
  geom_boxplot()
```


### Use `x = var_name_1` and `facet_grid(. ~ var_name_2)`

```{r}
cancer_clean %>%
  ggplot(aes(x = trt,
             y = weighin)) +
  geom_boxplot() + 
  facet_grid(. ~ stage)
```

### Use `fill = var_name_1` and `facet_grid(. ~ var_name_2)`

```{r}
cancer_clean %>%
  ggplot(aes(x = "Full Sample",
             y = weighin,
             fill = trt)) +
  geom_boxplot() + 
  facet_grid(. ~ stage)
```




## Multiple Boxes -for- a Subset AND -by- 2 Factors

The layers maybe combined to create more complicated plots.

Here is a plot for only participants who were in stage 1, 2, or 4 (n > 1 per stage), and compares the final oral condition for each stage between participants receiving the Aloe treatment vs. the placebo.


```{r}
cancer_clean %>%
  dplyr::filter(stage %in% c("1", "2", "4")) %>% 
  ggplot(aes(x = stage, 
             y = totalcw6,
             fill = trt)) +
  geom_boxplot() 
```



## Multiple Boxes -for- Repeated Measurements

The ordinal data is in what is called **"WIDE" format**, with the repeated measurements as separate variables, sitting side-by-side.  Each participant's data is contained in a single line *(25 lines here)*.

> To sort a dataset, use the `dplyr::arrange()` function and include the variable or variables you wish to sort by in the parentheses.

```{r}
cancer_clean %>% 
  dplyr::arrange(id) %>% 
  dplyr::select(id, totalcin, totalcw2, totalcw4, totalcw6)
```


Some data manipulations is needed to "stack" the repeated measurement variables (baseline, week 2, week 4, week 6) into a single variable we will call `value`.  We also need another new variable that denotes the measurement time, which we will call `time`.  The resulting dataset is said to be in **"LONG" format**.  Now there will be one line for each observation time on each participant *(4 x 25 = 100 lines here)*.

This is done with with the `tidyr::gather(key = "new_time_var", value = "new_value_var", old_var_1, old_var_2, ...)` function.

> Ignore the message that says *"attributes are not identical across measure variables; they will be dropped"*.

```{r}
cancer_clean %>%
  tidyr::pivot_longer(cols = c(totalcin, 
                               totalcw2, 
                               totalcw4, 
                               totalcw6),
                      names = "week",
                      names_pattern = "totalc(.)",
                      value = "totalc") %>% 
  dplyr::arrange(id, time) %>% 
  dplyr::select(id, time, value)
```

Once the data is in this format, you can create box plots for each time point.

> Ignore the additional message that says *"Removed 2 rows containing non-finite values (stat_boxplot)"*.  This is just alerting you to the fact that two people are missing their week 6 oral condiditon values.

```{r}
cancer_clean %>%
  tidyr::gather(key = "time", 
                value = "value", 
                totalcin, totalcw2, totalcw4, totalcw6) %>%
  ggplot(aes(x = time, 
             y = value)) +
  geom_boxplot()
```


### Multiple Boxes -for- Repeated Measurements AND -by- a Factor 

```{r}
cancer_clean %>%
  tidyr::gather(key = "time", 
                value = "value", 
                totalcin, totalcw2, totalcw4, totalcw6) %>%
  ggplot(aes(x = time, 
             y = value,
             fill = trt)) +
  geom_boxplot()
```


```{r}
cancer_clean %>%
  tidyr::gather(key = "time", 
                value = "value", 
                totalcin, totalcw2, totalcw4, totalcw6) %>%
  ggplot(aes(x = time, 
             y = value)) +
  geom_boxplot() +
  facet_grid(. ~ trt)
```




