# Binomial Sign Test

```{r}
library(tidyverse)
library(pander)
```

## The Binomial Distribution

```{r, fig.width=4, fig.height=4}
data.frame(count = 0:13) %>% 
  dplyr::mutate(prob = dbinom(count, size = 13, p = .50)) %>% 
  ggplot(aes(x = count,
             y = prob)) +
  geom_point() +
  geom_segment(aes(xend = count, 
                   y = 0, 
                   yend = prob),
               size = .1) +
  theme_bw() +
  scale_x_continuous() +
  labs(x = "Count",
       y = "Probability") +
  theme(legend.position = "none")
```

### Effect of Hypothesised "P" Value


```{r, fig.width=7, fig.height=3}
ex_n = 30

data.frame(count = 0:35) %>% 
  dplyr::mutate(prob_10 = dbinom(count, size = ex_n, p = .10)) %>% 
  dplyr::mutate(prob_25 = dbinom(count, size = ex_n, p = .25)) %>% 
  dplyr::mutate(prob_50 = dbinom(count, size = ex_n, p = .50)) %>% 
  dplyr::mutate(prob_75 = dbinom(count, size = ex_n, p = .75)) %>% 
  dplyr::mutate(prob_90 = dbinom(count, size = ex_n, p = .90)) %>%
  tidyr::pivot_longer(cols = contains("prob"),
                      names_to = "percent",
                      names_prefix = "prob_",
                      values_to = "prob") %>% 
  dplyr::mutate(percent = paste0("P = .", percent)) %>% 
  ggplot(aes(x = count,
             y = prob)) +
  geom_point() +
  geom_segment(aes(xend = count, 
                   y = 0, 
                   yend = prob),
               size = .1) +
  theme_bw() +
  scale_x_continuous() +
  labs(x = "Count",
       y = "Probability") +
  theme(legend.position = "none") +
  facet_grid(~ percent)
```


### Effect of Sample Size


### Probability of 4 *(or more)* out of 5


```{r}
ex_n = 5
ex_x = 4
ex_p = .50

dbinom(x    = ex_x, 
       size = ex_n, 
       prob = ex_p)
````




```{r, fig.width=3, fig.height=3}
data.frame(count = 0:ex_n) %>% 
  dplyr::mutate(prob = dbinom(count, size = ex_n, p = ex_p)) %>% 
  dplyr::mutate(mark = cut(count, 
                           breaks = c(0, ex_x - .5, ex_n),
                           include.lowest = TRUE)) %>% 
  ggplot(aes(x = count,
             y = prob,
             color = mark)) +
  geom_point(size = 2) +
  geom_segment(aes(xend = count, 
                   y = 0, 
                   yend = prob),
               size = 1) +
  theme_bw() +
  scale_x_continuous(breaks = 0:ex_n) +
  scale_color_manual(values = c("black", "red")) +
  labs(x = "Count",
       y = "Probability") +
  theme(legend.position = "none")
```


### Probability of 8 *(or more)* out of 10 

```{r}
ex_n = 10
ex_x = 8
ex_p = .50

dbinom(x    = ex_x, 
       size = ex_n, 
       prob = ex_p)
````


```{r, fig.width=3, fig.height=3}
data.frame(count = 0:ex_n) %>% 
  dplyr::mutate(prob = dbinom(count, size = ex_n, p = ex_p)) %>% 
  dplyr::mutate(mark = cut(count, 
                           breaks = c(0, ex_x -.5, ex_n),
                           include.lowest = TRUE)) %>% 
  ggplot(aes(x = count,
             y = prob,
             color = mark)) +
  geom_point(size = 2) +
  geom_segment(aes(xend = count, 
                   y = 0, 
                   yend = prob),
               size = 1) +
  theme_bw() +
  scale_x_continuous(breaks = 0:ex_n) +
  scale_color_manual(values = c("black", "red")) +
  labs(x = "Count",
       y = "Probability") +
  theme(legend.position = "none")
```

### Probability of 80 *(or more)* out of 100 

```{r}
ex_n = 100
ex_x = 80
ex_p = .50

dbinom(x    = ex_x, 
       size = ex_n, 
       prob = ex_p)
````

```{r, fig.width=3, fig.height=3}
data.frame(count = 0:ex_n) %>% 
  dplyr::mutate(prob = dbinom(count, size = ex_n, p = ex_p)) %>% 
  dplyr::mutate(mark = cut(count, 
                           breaks = c(0, ex_x, ex_n),
                           include.lowest = TRUE)) %>% 
  ggplot(aes(x = count,
             y = prob,
             color = mark)) +
  geom_point() +
  geom_segment(aes(xend = count, 
                   y = 0, 
                   yend = prob)) +
  theme_bw() +
  scale_x_continuous() +
  scale_color_manual(values = c("black", "red")) +
  labs(x = "Count",
       y = "Probability") +
  theme(legend.position = "none")
```

## Example: Fair Coin?

**Experiment**
A coin is flipped ten times and lands on heads eight times

**Quesiton**
Is coin biased?

**Hypotheses**
$$
H_0:  Proportion (X) = .50 \text{ in population} \\

H_1:  Proportion (X) \ne .50 \text{ in population} 
$$

### Enter tabulated data (subtotals)

```{r}
tab_coin <- c(heads = 8,
              tails = 2) %>%
  as.table()

tab_coin %>% 
  addmargins()
```

### Two-sided Test

**defaults**

* Fifty-fifty change: $H_0: P = .50 \rightarrow$ `p = 0.50`
* Two-sided alternative: `alternative = "two.sided"`
* 95% confidence interval: `conf.level = 0.95`

```{r}
tab_coin %>%
  binom.test()
```

### One-sided Test

```{r}
tab_coin %>%
  binom.test(alternative = "greater")
```


